On Adding Pattern Matching to Haskell-based Deeply Embedded Domain Specific Languages\\ \large{Not For Distribution}

### 2020-07-30 paper/paper.tex
Wording
Work on some technical details of 'isEven' tail recursion example

Note that the above pattern matches can be written as a single nested pattern
match. Both forms compile down to the same Core representation.

@@

\ttt{E (ProdMatch x y)} is equivalent to a curried function from \ttt{x} to
\ttt{y} in the expression language .  Note that \ttt{s} is a (potentially
nested) pair type. For example, \ttt{E (ProdMatch (a, (b, c)) r)} is
equivalent to \ttt{E (a -> E (b -> E (c -> E r)))}.

@@

\begin{lstlisting}
isEven :: Int -> Bool
isEven x =
    case x == 0 of
      True -> True
      False ->
        case x == 1 of
          True -> False
          False -> isEven (x - 2)
\end{lstlisting}

We translate this to the expression language, and then apply two of
the constructors we have defined.  The first of these is an initial 
\ttt{DoneExp} which we then wrap with a \ttt{TailRec} to provide the
representation of the iteration execution:

\begin{lstlisting}
isEven' :: E Int -> E (Int -> Bool)
isEven' x =
  TailRec ( DoneExp (
    IfThenElseExp (Equal x (rep 0))
      TrueExp
      (IfThenElseExp (Equal x (rep 1))
        FalseExp
        (isEven' (Sub x (rep 2))))
\end{lstlisting}

The transformation then proceeds by pushing the DoneExp constructor through
the conditionals, and the recursive call is transformed into a StepExp constructor.
This leaves us the the transformed example:

\begin{lstlisting}
isEven'' :: E Int -> E (Int -> Bool)
isEven'' x =
  TailRec (
    IfThenElseExp (Equal x (rep 0))
      (DoneExp TrueExp)
      (IfThenElseExp (Equal x (rep 1))
        (DoneExp FalseExp)
        (StepExp (Sub x (rep 2))))
\end{lstlisting}

>>>>

The above pattern matches could also be written as a single nested pattern
match. Both forms compile down to the same Core representation.

@@

\ttt{E (ProdMatch x y)} is equivalent to a curried function from \ttt{x} to
\ttt{y} in the expression language.  \ttt{s} is a (potentially
nested) pair type. For example, \ttt{E (ProdMatch (a, (b, c)) r)} is
equivalent to \ttt{E (a -> E (b -> E (c -> E r)))}.

@@

\begin{lstlisting}
isEven :: Int -> Bool
isEven x =
  internalize (externalize (
    case x == 0 of
      True -> True
      False ->
        case x == 1 of
          True -> False
          False -> isEven (x - 2)))
\end{lstlisting}

% We translate this to the expression language, and then apply two of
% the constructors we have defined.  The first of these is an initial 
% \ttt{Done} which we then wrap with a \ttt{runIter} to provide the
% representation of the iteration execution.
% The transformation then proceeds by pushing the DoneExp constructor through
% the conditionals, and the recursive call is transformed into a StepExp constructor.
% This leaves us the the transformed example:

We first translate this to use the \verb|Iter| type. The lambda is brought
inside of the \ttt{internalize (externalize ...)} call and wrapped in a call to
\ttt{runIter}. This inner expression is then eta expanded. Finally, each \verb|case|
branch which does not contain a recursive call is wrapped in a \verb|Done| and
every recursive call is substitute for an application of the \verb|Step| constructor.

In the following code, \verb|eta| is a fresh variable.

\begin{lstlisting}
isEven' :: Int -> Bool
isEven' =
  internalize (externalize
    (\ eta ->
      runIter (
        \ x ->
          case x == 0 of
            True -> Done True
            False ->
              case x == 1 of
                True -> Done False
                False -> Step (x - 2))
        eta))
\end{lstlisting}

After this initial tail recursion transformation, the rest of the plugin brings this
result into the expression language. \verb|TailRec| is substituted for \verb|runIter|,
\verb|DoneExp| is substituted for \verb|Done| and \verb|StepExp| is substituted for \verb|Step|.
Next, the pattern matching transformation occurs, producing the final result:

\break

\begin{lstlisting}
isEven'' :: Int -> Bool
isEven'' =
  internalize (
    (Lam (Var 1)
      (App
        (TailRec (Lam (Var 2)
          (IfThenElseExp (Equal (Var 2)
                                (rep 0))
            (DoneExp TrueExp)
            (IfThenElseExp (Equal (Var 1)
                                  (rep 1))
              (DoneExp FalseExp)
              (StepExp (Sub (Var 2)
                            (rep 2)))))))
        (Var 1))))
\end{lstlisting}
